Brother, Can You Spare a Manufacturing Job? How Voters React to Deindustrialization

### 2021-12-10 Final Draft/Final-Draft.tex
adding figures and text

@@

\author{
  Catherine Darin\\
  Zagreb Mukerjee\thanks{
  Darin: Harvard University, \href{mailto:cdarin@hks.harvard.edu}{cdarin@hks.harvard.edu}.
  Mukerjee: Harvard University, \href{mailto:zagrebmukerjee@fas.harvard.edu}{zagrebmukerjee@fas.harvard.edu}.
}
}

@@

This paper analyzes the relationship between deindustrialization, race,
and the election of Donald Trump in 2016. Previous analysis has
frequently focused on racial animus, trade, or the conjunction of the
two (\cite{Autor20}; \cite{che16}; \cite{BR21}). We extend the analysis
of \cite{Baccini21}, one of the first papers to examine the political
effects of deindustrialization while considering race and localization.
Deindustrialization - the transition of the U.S. economy away from
manufacturing - has led to widespread job loss and created adverse
knock-on effects for a wide swath of Americans. This experience has not
been uniform; rather, it has been concentrated in certain areas of the
country that were previously economically reliant on manufacturing
XXXXXXXXXXXX MAP FIGURE XXXXXXXXXXX. We believe the localized nature of
this analysis highlights the uneven and concentrated harms of
deindustrialization.

@@

Our analysis extends \cite{Baccini21} in several ways. First, we change
focus from a measure of gross losses in manufacturing jobs to net
losses. As discussed in \ref{datasec}, we believe this to be a more
accurate measure of deindustrialization. Second, we change the temporal
focus. Rather than focusing on manufacturing job loss in 2012-2015, we
extend our analysis back to 2004. This is an important change. As shown
XXXXXXXXXXXXXX FIGURE XXXXXXXXXXXXXXX, the bulk of U.S. manufacturing
job loss took place prior to 2012; in fact, the 2012-2015 period was
characterized by a slight rebound in manufacturing jobs. During this
rebound, counties that experienced the most pronounced manufacturing job
losses in prior years were also likely to experience the most pronounced
recovery XXXXXXXXXXXXXXX FIGURE XXXXXXXXXXXXX. Because of this v-shaped
recovery, focusing only on 2012-2015 job losses/gains paints a distorted
picture. Finally, we augment the instrumental-variable approach with one
based on covariate balancing, which produces similar results.

@@

@@

The results can be found in XXXXXXXXXXXXX. The regression coefficient in
Model 4, for instance, means that a loss of \(10\) manufacturing jobs
per \(100\) workers in a county (a relatively extreme value) corresponds
to a loss of \(3.2 \%\) in Democratic vote share.

@@

Restricting our analysis to only the 2012-2015 period produces an
opposite effect, in contrast to \cite{BW21} (when we use a net job loss
metric). The loss of Democratic vote share seems to be associated with a
gain in manufacturing jobs, both in general and among whites. We believe
this time-restricted analysis suffers from an omitted variable bias. As
discussed above, the 2012-2015 period can be broadly characterized as a
slight rebound from the manufacturing job losses in the 2008 financial
crisis. The counties that experience job losses from 2004-2012 are more
likely to have job gains in 2012-2015. An apparent relationship between
2012-2015 job gains and loss of Democratic vote share in fact represents
2004-2012 job losses that may lead to both the later job gains and the
loss of vote share. XXXXXXXXXXXXX FIGURE XXXXXXXXXXXX

This result suggests an important temporal dimension to the political
effects of deindustrialization - job losses in 2004-2012 make their
effects felt in 2016. In some ways, this result is more plausible than
if the 2012-2015 job losses were a significant covariate of voting. The
effects of job loss may take some time to be felt, at both individual
and community levels (see XXXXXXXXXXXXXX CITE). The mechanisms of this
relationship make a good topic for subsequent research.

@@

As much research has noted, the national and global implications of
American deindustrialization are only beginning to be felt - from the
2016 election to isolationism to the opioid epidemic. Previous analyses
have often focused on choosing between explanations grounded in racial
issues or political-economic evolutions XXXXXXXXXXXXXXXXXXXXXxxx CITE.
We are joining a growing body of scholars searching for explanations at
the intersection of the two. We believe that deindustrialization is
fundamentally connected to particular places, and so the localized
approach adopted by \cite{BW21} offers a great deal of promise. Further
research might explore further ways in which geography has influenced
the rightward turn in the U.S., for example through the use of
techniques of spatial econometrics or the careful use of ethnography.

While our study is oriented towards history, deindustrialization is in
no way confined to the past. The trajectory of American manufacturing
employment is far from certain, with continued competition from abroad
and the perennial threat of unemployment. We believe the case of the
U.S. also offers insight that is applicable to the phenomenon of
`premature deindustrialization' in the developing world, and the
concurrent growth of populist parties in Hungary, Brazil, India and
elsewhere. XXXXXXXXXXXXXXX CITE XXXXXXxx

\begin{figure}

\caption{Caption for figure below.}
\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\begin{figurenotes}[Source]
Figure notes with optional leadin (Source, in this case).
\end{figurenotes}
\end{figure}

\nocite{stargazer}

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\author{
  Catherine Darin\\
  Zagreb Mukerjee\thanks{
  Darin: Harvard
University, \href{mailto:cdarin@hks.harvard.edu}{cdarin@hks.harvard.edu}.
  Mukerjee: Harvard
University, \href{mailto:zagrebmukerjee@fas.harvard.edu}{zagrebmukerjee@fas.harvard.edu}.
  Code available at https://github.com/zagrebmukerjee/ReplicationPaper
}
}

@@

This paper analyzes the relationship between deindustrialization, race,
and the election of Donald Trump in 2016. Previous analysis has
frequently focused on racial animus, trade, or the conjunction of the
two (\cite{Autor20}; \cite{Che16}; \cite{BR21}). We extend the analysis
of \cite{Baccini21}, one of the first papers to examine the political
effects of deindustrialization while considering race and localization.
Deindustrialization - the transition of the U.S. economy away from
manufacturing - has led to widespread job loss and created adverse
knock-on effects for a wide swath of Americans. This incorporates job
losses due to trade with China and others (see \cite{Acemoglu16}); it
also includes job losses due to automation, demand changes, and other
factors. The experience of deindustrialization has not been uniform;
rather, it has been concentrated in certain areas of the country that
were previously economically reliant on manufacturing XXXXXXXXXXXX MAP
FIGURE XXXXXXXXXXX. We believe the localized nature of this analysis
highlights the uneven and concentrated harms of deindustrialization.

@@

Our analysis extends \cite{Baccini21} in several ways. First, we change
focus from a measure of gross losses in manufacturing jobs to net
losses. As discussed in \ref{datasec}, we believe this to be a more
accurate measure of deindustrialization. Second, we change the temporal
focus. Rather than focusing on manufacturing job loss in 2012-2015, we
extend our analysis back to 2004. This is an important change. As shown
in Figure \ref{natlPlot}, the bulk of U.S. manufacturing job loss took
place prior to 2012; in fact, the 2012-2015 period was characterized by
a slight rebound in manufacturing jobs. During this rebound, counties
that experienced the most pronounced manufacturing job losses in prior
years were also likely to experience the most pronounced recovery, as
described in Figure \ref{vShapePlot}. Because of this v-shaped recovery,
focusing only on 2012-2015 job losses/gains paints a distorted picture.
Finally, we augment the instrumental-variable approach with one based on
covariate balancing, which produces similar results.

\begin{figure} \label{natlPlot}
\caption{Nationwide Change in Manufacturing Employment, 1995-2019}

\begin{center}\includegraphics{Final-Draft_files/figure-latex/unnamed-chunk-2-1} \end{center}

\begin{figurenotes}
Gain or loss of national manufacturing jobs, divided by total national workforce.
\end{figurenotes}
\end{figure}

\begin{figure} \label{vShapePlot}
\caption{}

\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\end{figure}

@@

Basic descriptive statistics of job losses and other key covariates can
be seen in Table \ref{tab:descTable}.

\begin{table}[!h]

\caption{\label{tab:descTable}Manufacturing Job Changes 2004-2015}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrr}
\toprule
  & Mean & Std Dev & 25th Pctile & Median & 75th Pctile\\
\midrule
Mfg Share of Emp & 0.20 & 0.15 & 0.09 & 0.17 & 0.29\\
Mfg Share of Emp (White) & 0.16 & 0.13 & 0.06 & 0.13 & 0.22\\
Mfg Share of Emp (Nonwhite) & 0.05 & 0.07 & 0.01 & 0.02 & 0.06\\
Change in Mfg Jobs/ Worker & 0.04 & 0.09 & -0.01 & 0.02 & 0.06\\
Change in Mfg Jobs/ Worker (W) & 0.03 & 0.07 & 0.00 & 0.02 & 0.06\\
\addlinespace
Change in Mfg Jobs/ Worker (NW) & 0.00 & 0.04 & -0.01 & 0.00 & 0.01\\
\bottomrule
\end{tabular}}
\end{table}

@@

The results can be found in Table \ref{regResult04}. The regression
coefficient in Model 4, for instance, means that a loss of \(10\)
manufacturing jobs per \(100\) workers in a county (a relatively extreme
value) corresponds to a loss of \(3.2 \%\) in Democratic vote share.

\begin{table}[!htbp] \centering 
  \caption{Effect of Manufacturing Job Loss from 2004 to 2015} 
  \label{regResult04} 
\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Change in Dem. Share, 2016-2012} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Manufacturing Layoffs & $-$0.34$^{**}$ & $-$0.38$^{**}$ &  &  \\ 
  & (0.14) & (0.17) &  &  \\ 
  & & & & \\ 
 White Manufacturing Layoffs &  &  & $-$0.23$^{***}$ & $-$0.32$^{***}$ \\ 
  &  &  & (0.08) & (0.10) \\ 
  & & & & \\ 
 Nonwhite Manufacturing Layoffs &  &  & 0.15$^{***}$ & 0.20$^{***}$ \\ 
  &  &  & (0.05) & (0.06) \\ 
  & & & & \\ 
\hline \\[-1.8ex] 
White Share/Svc Layoffs Control & No & Yes & No & Yes \\ 
Observations & 3,049 & 3,049 & 2,750 & 2,750 \\ 
Adjusted R$^{2}$ & 0.72 & 0.72 & 0.75 & 0.75 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

@@

Restricting our analysis to only the 2012-2015 period, while continuing
to use the net job loss metric, produces an effect opposite to that of
\cite{Baccini21}. The loss of Democratic vote share seems to be
associated with a gain in manufacturing jobs, both in general and among
whites. We believe this time-restricted analysis suffers from an omitted
variable bias. As discussed above, the 2012-2015 period can be broadly
characterized as a slight rebound from the manufacturing job losses in
the 2008 financial crisis. The counties that experience job losses from
2004-2012 are more likely to have job gains in 2012-2015 (see
\ref{vShapePlot}). An apparent relationship between 2012-2015 job gains
and loss of Democratic vote share in fact represents 2004-2012 job
losses that may lead to both the later job gains and the loss of vote
share. These results can be seen in Table \ref{regResult12}.

\begin{table}[!htbp] \centering 
  \caption{Effect of Manufacturing Job Loss from 2004 to 2015} 
  \label{regResult12} 
\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Change in Dem. Share, 2016-2012} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Manufacturing Layoffs & 0.43$^{**}$ & 0.37$^{**}$ &  &  \\ 
  & (0.17) & (0.17) &  &  \\ 
  & & & & \\ 
 White Manufacturing Layoffs &  &  & 6.50$^{***}$ & 2.33$^{***}$ \\ 
  &  &  & (2.20) & (0.76) \\ 
  & & & & \\ 
 Nonwhite Manufacturing Layoffs &  &  & $-$3.99$^{***}$ & $-$1.42$^{***}$ \\ 
  &  &  & (1.36) & (0.47) \\ 
  & & & & \\ 
\hline \\[-1.8ex] 
White Share/Svc Layoffs Control & No & Yes & No & Yes \\ 
Observations & 3,064 & 3,064 & 2,765 & 2,765 \\ 
Adjusted R$^{2}$ & 0.73 & 0.73 & 0.75 & 0.75 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

This result suggests an important temporal dimension to the political
effects of deindustrialization - job losses in 2004-2012 make their
effects felt in 2016. In some ways, this result is more plausible than
if the 2012-2015 job losses were a significant covariate of voting. The
effects of job loss may take some time to be felt, at both individual
and community levels (see \cite{Mckee09}, \cite{Foote19}). The
mechanisms of this relationship make a good topic for subsequent
research.

@@

As much research has noted, the national and global implications of
American deindustrialization are only beginning to be felt - from the
2016 election to isolationism to the opioid epidemic. Previous analyses
have often focused on choosing between explanations grounded in racial
issues or political-economic evolutions (ex: \cite{Green19},
\cite{Reny19}). We are joining a growing body of scholars searching for
explanations at the intersection of the two. We believe that
deindustrialization is fundamentally connected to particular places, and
so the localized approach adopted by \cite{Baccini21} offers a great
deal of promise. Further research might explore further ways in which
geography has influenced the rightward turn in the U.S., for example
through the use of techniques of spatial econometrics or the careful use
of ethnography.

While our study is oriented towards history, deindustrialization is in
no way confined to the past. The trajectory of American manufacturing
employment is far from certain, with continued competition from abroad
and the perennial threat of automation. We believe the case of the U.S.
also offers insight that is applicable to the phenomenon of `premature
deindustrialization' in the developing world, and the concurrent growth
of populist parties in Hungary, Brazil, India and elsewhere
(\cite{Rodrik15}, \cite{Castillo16}).

\begin{figure} \label{}
\caption{Caption for figure below.}

\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\begin{figurenotes}[Source]
Figure notes with optional leadin (Source, in this case).
\end{figurenotes}
\end{figure}

\nocite{Stargazer}
